Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-03-12 paper/manuscript.tex
Some minor fixes  On branch sboresch  Changes to be committed: 	modified:   manuscript.tex 	modified:   sb.bib

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems; in particular, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see {\bf XXX}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches in practical use differ. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation hardware.

@@

When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules. %Often this is predetermined by the choice of simulation software.
As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}. The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state. 
Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.

@@

\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.

@@

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{Ssee below; where do we place the dummy atom discussion?}. Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

@@

Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~{\bfseries XX}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters. E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$. For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond. The other energy terms are treated analogously. The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.  More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.

>>>>

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see {\bf XXX}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.

@@

When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules. %Often this is predetermined by the choice of simulation software.
As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}. The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}. \sbnote{Fig.~3 needs to be updated for dual coordinate methods and corrected, some bonds for the dummy atoms have the wrong anchor atom!}
In this case the common substructure between the two molecules is the benzene ring, though in practice the substructure may be selected to be larger depending on the mapping chosen, as we discuss below.

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by first changing one of the hydrogens into a carbon. The site of this transformation will also be the future home of two additional hydrogen atoms bound to the new carbon, so these must initially be present as non-interacting atoms called ``dummy atoms'', which retain their bonded interactions but do not interact with the rest of the system.  Bond parameters as well as partial charges between the changing atoms are adjusted accordingly between the initial and final state; see below for further details. Thus, in a single topology calculation, atoms may change their type, ensuring minimal dummy atoms are created. This is illustrated in the left arm of Fig.~\ref{fig:fig_topology}.

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other.

@@

\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional. For an in-depth analyis of differences resulting from the use of a  single vs.\ a dual topology approach (single coordinate in both cases) see Ref.~\cite{Boresch_2002}.

@@

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{See below; where do we place the dummy atom discussion?}. Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

@@

Dual coordinate approaches with chemical topologies encompass a third group of methods, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space. Recently, a variant of $\lambda$-dynamics that uses chemical topologies (separate coordinates for each state) has been proposed \cite{Liesen_2024}.

@@

Concerning SC approaches, Pearlman’s usage of the term single topology not only meant a chimeric construct as depicted in Fig.~\ref{fig:fig_topology}, but specifically implied that the coupling parameter $\lambda$ was applied to the force field parameters. E.g., a bond stretching energy term changing from C-H to C-C has the functional form $U_{bond}(\lambda)=K(\lambda)(r-r_0(\lambda))^2$. For $\lambda=0$, $K(\lambda)$ and $r_0(\lambda)$ correspond to the values appropriate for an aliphatic C-H bond, and for $\lambda=1$ to those for an aliphatic C-C bond. The other energy terms are treated analogously. The single topology paradigm as just summarized was the default mode of alchemical transformations in AMBER and GROMOS at that time.  More recently, this implementation of single topology has been referred to as parameter interpolation or mixing \cite{Giese_2018}.
